Read Me

### 2023-04-25 code/analysis_code/1_EDA_alt.Rmd
more changes

```{r message=FALSE, warning=FALSE}
library(here)
library(spdep)
library(GGally)
library(skimr)
library(ggfortify)
library(MASS)
library(ggcorrplot)
library("FactoMineR")
library(tidyverse)
library(broom)
library(knitr)
library(ggpubr)
```

@@

@@

Let's start by looking at the overall HIV-related death rate by race in the United States.
```{r}
race_sub %>% 
  group_by(Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  mutate(percent_by_race = round(tot_deaths/sum(tot_deaths)*100, 1),
         ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population (>12 years old)",
                      "Death Rate (per 100,000)",
                      "Percent",
                      "Ratio"))
# using white persons as reference group
```

@@

```

Let's move on to
Bivariate Analysis
```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Living in Poverty`)) +
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Living in Poverty") + 
  stat_cor(method = "pearson") 
#  pass
ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
```

@@

```{r}
ggplot(race_sub, aes(y = MRR, x = `Gini Coefficient`)) +
    geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Gini Coefficient") +
  stat_cor(method = "pearson") 
ggsave(here("results/bi_gini.png"), units="in", width=10, height=7, dpi=300)
# did not "pass"
```

@@

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Without an Internet subscription`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Without an Internet Subscription") + 
  stat_cor(method="pearson")
# "pass"
```

@@

```{r}
ggplot(race_sub, aes(y = MRR, x = `Median Household Income`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Median Household Income") + 
  stat_cor(method="pearson")
# "pass"
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percentage Diagnosed Late`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percentage Diagnosed Late") + 
  stat_cor(method="pearson")
# "pass"
ggsave(here("results/bi_latedx.png"), units="in", width=10, height=7, dpi=300)
```

@@

```{r}
pos_predictors <- race_sub %>% 
  select(c("State", 
           "Race",
           "Percent Living in Poverty",
           "Percent Without an Internet subscription",
           "Percent Uninsured Under 65 Years Old",
           "Percent Living with Food Insecurity",
           "Percent Living with Severe Housing Cost Burden",
           "Median Household Income",
           "Percentage Diagnosed Late"))
```

>>>>

```{r message=FALSE, warning=FALSE}
library(here)
library(spdep)
library(GGally)
library(skimr)
library(ggfortify)
library(MASS)
library(ggcorrplot)
library("FactoMineR")
library(tidyverse)
library(broom)
library(knitr)
library(kableExtra)
library(ggpubr)
```

@@

```{r}
states <- data.frame(state.name)
states$Region <- case_when(
  states$state.name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania") ~ "Northeast",
  states$state.name %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest",
  states$state.name %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South",
  states$state.name %in% c("Arizona", "California", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West"
)

race_sub <- full_join(race_sub, states, by = join_by(State == state.name))

race_sub <- race_sub %>% 
  filter(!grepl("Alaska", State),
         !grepl("Hawaii", State))
```

@@

Let's start by looking at the overall HIV-related death rate by race in the United States.
```{r}
table1 <- race_sub %>% 
  group_by(Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population \n (>12 years old)",
                      "Death Rate \n (per 100,000)",
                      "Mortality Rate Ratio"),
        caption = "<center>HIV-Related Deaths by Race/ Ethnicity, United States, 2020</center>",
        align = "c") %>%
  kable_minimal() %>%
  kable_styling(font_size = 30)%>%
gsub("font-size: initial !important;",
"font-size: 30pt !important;",
.)  %>% 
  landscape()

table1

table1 %>%  
  save_kable(here("results/table1.png"))
# using white persons as reference group
```

```{r}
region_table <- race_sub %>% 
  group_by(Region, Race) %>% 
  summarize(tot_deaths = sum(`Death Cases`),
            tot_population = sum(`Population`),
            overall_rate = round(tot_deaths/ tot_population *1e5,1)) %>% 
  ungroup() %>% 
  group_by(Region) %>% 
  mutate(ratio = round(overall_rate/overall_rate[Race =="White"],1)) %>%
  kable(col.names = c("Region",
                      "Race/ Ethnicity",
                      "HIV-Related Deaths (n)",
                      "Population \n (>12 years old)",
                      "Death Rate \n (per 100,000)",
                      "Mortality Rate Ratio"),
        caption = "<center>HIV-Related Deaths by Region, United States, 2020</center>") %>% 
  kable_minimal() %>%
  kable_styling(font_size = 30) %>%
gsub("font-size: initial !important;",
"font-size: 30pt !important;",.)  %>% 
  landscape()

region_table

region_table %>%  
  save_kable(here("results/region_table.png"))
# using white persons as reference group
```

@@

race_sub %>% 
  ggplot() +
  geom_boxplot(aes(y = `Death Rate (estimated)`, x = `Region`, group = `Region`)) + theme_classic() + 
  theme(legend.position = "bottom") +
  labs(title = "HIV-Related Death Rate of Black and White Persons, by Region, United States, 2020", y = "Rate (per 100,000)", x = "Region") + facet_wrap(.~ Race)
ggsave(here("results/dr_region.png"), units="in", width=12, height=5, dpi=300)
```

Let's move on to
Bivariate Analysis 
We will deem variables with a correlation coeffiencient of |0.15| as important, and will include these in PCA
```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Living in Poverty`)) +
  geom_smooth(method = "lm") + 
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Living in Poverty") + 
  stat_cor(method = "pearson") 
#  pass 
ggsave(here("results/bi_pov.png"), units="in", width=10, height=7, dpi=300)
```

@@

```{r}
ggplot(race_sub, aes(y = MRR, x = `Gini Coefficient`)) +
    geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Gini Coefficient") +
  stat_cor(method = "pearson") 
#"pass"
ggsave(here("results/bi_gini.png"), units="in", width=10, height=7, dpi=300)
```

@@

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percent Without an Internet subscription`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percent Without an Internet Subscription") + 
  stat_cor(method="pearson")
# "pass"
ggsave(here("results/bi_no_internet.png"), units="in", width=10, height=7, dpi=300)
```

@@

```{r}
ggplot(race_sub, aes(y = MRR, x = `Median Household Income`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Median Household Income") + 
  stat_cor(method="pearson")
# did not "pass"
```

```{r}
ggplot(race_sub, aes(y = MRR, x = `Percentage Diagnosed Late`)) +
  geom_smooth(method = "lm") +
  geom_point() + 
  labs(title = "Black-to-White HIV-Related Mortality Rate Ratio (MRR) vs.
       Percentage Diagnosed Late") + 
  stat_cor(method="pearson")
# did not"pass"
```

@@

```{r}
pos_predictors <- race_sub %>% 
  select(c("State", 
           "Race",
           "Gini Coefficient",
           "Percent Living in Poverty",
           "Percent Without an Internet subscription",
           "Percent Uninsured Under 65 Years Old",
           "Percent Living with Food Insecurity"))
```

### 2023-04-25 code/analysis_code/2_pca_alt.Rmd
more changes

  #separating the y variables
df_response <- race_sub %>%
    select(contains(c("State", "Ratio", "MRR", "Race", "Death", "Population")))

  # pulling variables that were correlated with MRR, and thenp scaling
pos_predictors_scaled <- pos_predictors %>% 
    ungroup() %>% 
    mutate(across(where(is.numeric), scale))

#Setting up correlation matrix
corr_matrix <- pos_predictors_scaled %>% 
    select(c(3:9)) %>% 
    cor()

@@

#pca loadings of principal components 1 and 2
print(pca$loadings[, 1:2])

#Scree plot
print(fviz_eig(pca, addlabels = TRUE))
ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
```
Since we originally had 11 variables, we should only look at components that explain more than 9.1% of variation. We can see from the scree plot that principal component 1 explained 73.3% of variation, and components 1 and 2 explained 94.9% of the variation, and components 1 - 2 explained 91.1%. All other principal components explains less that 8.9% of variation, so they will excluded.

>>>>

  #separating the y variables
df_response <- race_sub %>%
    select(contains(c("State","Region", "Ratio", "MRR", "Race", "Death", "Population")))

  # pulling variables that were correlated with MRR, and then scaling
pos_predictors_scaled <- pos_predictors %>% 
    ungroup() %>% 
    mutate(across(where(is.numeric), scale))

#Setting up correlation matrix
corr_matrix <- pos_predictors_scaled %>% 
    select(c(3:7)) %>% 
    cor()

@@

#pca loadings of principal components 1 and 2
print(pca$loadings[, 1:3])

#Scree plot
print(fviz_eig(pca, addlabels = TRUE))
ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
```
Since we originally had 5 variables, we should only look at components that explain more than 20% of variation. This is includes components 1 through 3. All other principal components explains less that 8.9% of variation, so they will excluded.

### 2023-04-25 code/analysis_code/3_model_alt.Rmd
more changes

Setting up functions for models
```{r}
 #multiple linear regression
mult_function <- function(df) {
  mult_race <- glm(MRR ~ 
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                     #   `Percent Living with Severe Housing Cost Burden` +
                        `Percent Living with Food Insecurity`, 
                     data = df,
                     family = "gaussian")
  print(summary(mult_race))
    ##pull fitted values
df$mult_fit <- mult_race$fitted.values
return(df)
}

## function to fit GLM and pull AIC
mult_aic <- function(df) {
  mult_race <- glm(MRR ~ 
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                       # `Percent Living with Severe Housing Cost Burden` +
                        `Percent Living with Food Insecurity`,
                     data = df,
                     family = "gaussian")
aic <- AIC(mult_race)

@@

pois_race <- glm(`Death Cases` ~ Race +
                        `Percent Living in Poverty` +
                       # `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                        `Percent Living with Food Insecurity` +
                        #`Percent Living with Severe Housing Cost Burden` +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
  print(summary(pois_race))

@@

## function to fit GLM and pull AIC
pois_aic <- function(df) {
  pois_race <- glm(`Death Cases` ~  Race +
                        `Percent Living in Poverty` +
                      #  `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                        `Percent Living with Food Insecurity` +
                       # `Percent Living with Severe Housing Cost Burden` +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
aic <- AIC(pois_race)
return(aic)
}
```

@@

```{r}
  #Negative binomial
nb_function  <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race +
                        `Percent Living in Poverty` +
                       # `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                        `Percent Living with Food Insecurity`  +
                       # `Percent Living with Severe Housing Cost Burden` +
                     offset(log(`Population`)),
                            data = df)
  print(summary(neg_bin_race))

@@

## function to fit GLM and pull AIC
nb_aic <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race +
                        `Percent Living in Poverty` +
                       # `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                        `Median Household Income`  +
                        `Percent Living with Food Insecurity`  +
                        #`Percent Living with Severe Housing Cost Burden` +
                     offset(log(`Population`)),
                            data = df)
aic <- AIC(neg_bin_race)

@@

  car_race <- spautolm(MRR ~                      
                        `Percent Living in Poverty` +
                       `Median Household Income` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                     #   `Percent Unemployed`  +
                     #   `Percent Living with Severe Housing Cost Burden` +
                        `Percent Living with Food Insecurity`, 
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))

@@

  car_race <- spautolm(MRR ~                      
                        `Percent Living in Poverty` +
                       `Median Household Income` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                   #     `Percent Unemployed`  +
                      #  `Percent Living with Severe Housing Cost Burden` +
                        `Percent Living with Food Insecurity`, 
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
aic <- AIC(car_race)
return(aic)

@@

Organizing mod_AIC into new dataframe
```{r}
aic_df <- data.frame(AIC = c(df_mult, df_pois, df_nb, df_car),
                     Model = c("GLM", "Poisson", "Negative Binomial", "CAR"),
                     Outcome = c("Ratio", "Count", "Count", "Ratio"))
aic_df <- aic_df %>% mutate(AIC  = round(AIC, 2))
```

```{r}
saveRDS(df, here("data/processed_data/df_fit_alt.rds"))

saveRDS(aic_df, here("data/processed_data/aic_alt.rds"))
```

>>>>

Setting up functions for models
```{r}
 #multiple linear regression
mult_function <- function(df) {
  mult_race <- glm(MRR ~ 
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old`,
                      #  `Percent Living with Food Insecurity`,
                     data = df,
                     family = "gaussian")
  print(summary(mult_race))
    ##pull fitted values
df$mult_fit <- mult_race$fitted.values
return(df)
}

## function to fit GLM and pull AIC
mult_aic <- function(df) {
  mult_race <- glm(MRR ~ 
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old`,
                      #  `Percent Living with Food Insecurity`,
                     data = df,
                     family = "gaussian")
aic <- AIC(mult_race)

@@

pois_race <- glm(`Death Cases` ~ Race + region +
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                      #  `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
  print(summary(pois_race))

@@

## function to fit GLM and pull AIC
pois_aic <- function(df) {
  pois_race <- glm(`Death Cases` ~  Race + region +
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                      #  `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
aic <- AIC(pois_race)
return(aic)
}
```

@@

```{r}
  #Negative binomial
nb_function  <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                      #  `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                            data = df)
  print(summary(neg_bin_race))

@@

## function to fit GLM and pull AIC
nb_aic <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                      #  `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                            data = df)
aic <- AIC(neg_bin_race)

@@

  car_race <- spautolm(MRR ~                      
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old`,
                      #  `Percent Living with Food Insecurity`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))

@@

  car_race <- spautolm(MRR ~                      
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old`,
                      #  `Percent Living with Food Insecurity`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
aic <- AIC(car_race)
return(aic)

@@

Organizing mod_AIC into new dataframe
```{r}
aic_df <- data.frame(AIC = c(df_mult, df_pois, df_nb, df_car),
                     Model = c("Multiple Linear Regression", "Poisson", "Negative Binomial", "Conditional Autoregressive"),
                     Outcome = c("Ratio", "Count", "Count", "Ratio"))
aic_df <- aic_df %>% mutate(AIC  = round(AIC, 2))

```{r}
library(kableExtra)
library(knitr)
table_fit <- aic_df %>%
  select(c(Model, Outcome, AIC)) %>% 
  kable(caption = "<center>Model Fit</center>",
        align = "c") %>%
  kable_minimal() %>%
  kable_styling(font_size = 30) %>%
gsub("font-size: initial !important;",
"font-size: 30pt !important;",
.)  %>% 
  landscape()

table_fit %>%  
  save_kable(here("results/table_fit.png"))

```{r}
saveRDS(df, here("data/processed_data/df_fit_alt.rds"))
```

Creating function that subsets by specified race, and compute/ plots Moran's I and lisa.
```{r}
spacy_function <- function(df) {

  h_race_df <- matrix(0, nrow(df), nrow(df))
  h_race_df[cbind(match(us_nabs$StateCode, df$`State Abbreviation`),
  match(us_nabs$NeighborStateCode, df$`State Abbreviation`))] <- 1

 # plot Moran's I
  print(moran.test(df$MRR, nablist_race_df, zero.policy = TRUE, randomisation = FALSE))
  moran.plot(df$MRR, nablist_race_df, zero.policy = TRUE)

  # calculate and print LISA statistics
  lisa_race <- localmoran(df$MRR, nablist_race_df)
  summary(lisa_race[,1])
}
```

```{r}
spacy_function(df)
```

```{r}
png(here("results/moran_mrr.png"))
spacy_function(df)
dev.off()
```
